# Set up your environment

To build an R package you will need [R](https://cran.r-project.org/bin/windows/base/) installed and an IDE: preferably [RStudio](https://posit.co/download/rstudio-desktop/) or [Positron](https://github.com/posit-dev/positron/releases).

You will need to install [devtools](https://devtools.r-lib.org) and [usethis](https://usethis.r-lib.org) as they are key to follow the different steps to get started.

```{r, eval=FALSE}
# use base R
install.packages(c("usethis", "devtools"))

# or pak (recommended)
library(pak)
pkg_install(c("usethis", "devtools"))
```

To work with the Tidy R OMOP ecosystem, you will also need to install the [omopgenerics](https://darwin-eu.github.io/omopgenerics/) package. Alongside with other packages that may be useful:

-   [PatientProfiles](https://darwin-eu.github.io/PatientProfiles/) for patient-level feature extraction.
-   [CodelistGenerator](https://darwin-eu.github.io/CodelistGenerator/) to query the OMOP vocabularies.
-   [visOmopResults](https://darwin-eu.github.io/visOmopResults/) for formatting and visualising results.

You can see the full list of Tidy R packages in our [website](https://oxford-pharmacoepi.github.io/Oxinfer/packages/summary.html).

## Create an empty package

To create a new empty package, use `usethis::create_package()` with the path to the folder where you want the package to live. The folder name becomes the package name.

```{r, eval=FALSE}
library(usethis)

create_package(path = "path/to/MyPackage")
```

This will:

- Create the folder `MyPackage/` with the standard package structure (`R/`, `DESCRIPTION`, `NAMESPACE`).
- Open the new project in a fresh RStudio or Positron window.

::: {.callout-tip}
### Choosing a package name

Package names must start with a letter, contain only letters, numbers, and dots, and ideally be short, distinctive, and easy to remember. Use CamelCase to make multi-word names readable (e.g. `CohortConstructor`, `PatientProfiles`). Avoid dots, underscores, and all-lowercase names that are hard to distinguish.

You can check whether a name is available on CRAN with `available::available("MyPackageName")`.
:::

### The package structure

After `create_package()` you will have:

```
MyPackage/
├── R/                  # Your R source files go here
├── DESCRIPTION         # Package metadata (name, version, dependencies, …)
├── NAMESPACE           # Exported functions — managed automatically by devtools
└── MyPackage.Rproj     # RStudio project file
```

You will add more files as you go: `tests/`, `vignettes/`, `man/`, and `NEWS.md` among others. Most of these are created for you by `usethis` helpers described later in this chapter.

## Add a license

Adding a license is essential — it tells users how they can use, modify, and redistribute your package, and protects you legally. In the Tidy R OMOP ecosystem we release under the [Apache 2.0](https://www.apache.org/licenses/LICENSE-2.0) license.

```{r, eval=FALSE}
use_apache_license()
```

This adds the `LICENSE.md` file and populates the `License` field in `DESCRIPTION`.

::: {.callout-note collapse="true"}
### License options

`usethis` supports all common open-source licenses. The right choice usually comes down to how permissive you want to be:

-   More permissive (anyone can use your code, even in closed-source projects):
    -   [MIT](https://choosealicense.com/licenses/mit/): simple and widely understood.
    -   [Apache 2.0](https://choosealicense.com/licenses/apache-2.0/): like MIT but adds explicit patent protection. This is our default.

-   Copyleft (anyone who distributes changes must also open-source them):
    -   [GPL v3](https://choosealicense.com/licenses/gpl-3.0/)
    -   [AGPL v3](https://choosealicense.com/licenses/agpl-3.0/): extends GPL to cover use over a network.
    -   [LGPL v3](https://choosealicense.com/licenses/lgpl-3.0/): a weaker copyleft, suitable for libraries.

-   Creative Commons (appropriate for data packages rather than code):
    -   [CC0](https://creativecommons.org/publicdomain/zero/1.0/): dedicated to the public domain.
    -   [CC-BY](https://creativecommons.org/licenses/by/4.0/): free to share and adapt, with attribution.

```{r, eval=FALSE}
use_mit_license()
use_apache_license()
use_gpl_license(version = 3)
use_agpl_license(version = 3)
use_lgpl_license(version = 3)
use_cc0_license()
use_ccby_license()
```
:::

## Create a first function

All R source files live in the `R/` folder. Each file can contain one or more functions — the typical convention is one file per function or per closely related group of functions, named to match.

Create a new R file with:

```{r, eval=FALSE}
use_r("myFunction")
```

This creates `R/myFunction.R` and opens it. Write your function there:

```{r, eval=FALSE}
myFunction <- function(cohort, overlap = FALSE) {
  # your code here
  cohort
}
```

::: {.callout-tip}
### Add a roxygen skeleton immediately

Before writing the body, add a roxygen2 documentation skeleton with **Code → Insert Roxygen Skeleton** in RStudio (or `Ctrl+Alt+Shift+R`). Fill in the `@title`, `@description`, `@param`, and `@return` tags as you write the function. Doing this from the start is much easier than adding it later. Documentation is covered fully in @sec-functions.
:::

Once you have written a function, load the whole package into your session with:

```{r, eval=FALSE}
devtools::load_all()  # Ctrl+Shift+L in RStudio
```

`load_all()` simulates what happens when the package is installed and loaded — it makes all your functions available without having to restart R. Use it constantly during development.

## Check the package

`devtools::check()` runs `R CMD check`, which is the official test that CRAN and you should run to verify your package is correct. It checks documentation, dependencies, examples, and tests.

```{r, eval=FALSE}
devtools::check()  # Ctrl+Shift+E in RStudio
```

A clean check produces output ending with:

```
── R CMD check results ─────────────────────────────────────────────────
Duration: 30s

0 errors ✔ | 0 warnings ✔ | 0 notes ✔
```

Aim to keep your package at **0 errors, 0 warnings, 0 notes** at all times — not just before a release. It is much easier to fix one new note immediately than to diagnose five notes that accumulated over months.

Common things to fix early:

- A missing `@export` tag on a function that should be public.
- An undocumented argument (`@param` entry missing).
- A package used in code but not listed in `DESCRIPTION` under `Imports` or `Suggests`.
- An example that fails because it requires a real database.

## Create a simple test

Tests live in `tests/testthat/`. Set up the testing infrastructure with:

```{r, eval=FALSE}
use_testthat()
```

Then create a test file for your function:

```{r, eval=FALSE}
use_test("myFunction")
```

This creates `tests/testthat/test-myFunction.R`. A minimal test looks like:

```{r, eval=FALSE}
test_that("myFunction returns a cohort table", {
  cdm <- mockPatientProfiles()
  result <- myFunction(cdm$cohort1)
  expect_s3_class(result, "cohort_table")
  CDMConnector::cdmDisconnect(cdm)
})
```

Run the tests for the file you are currently editing:

```{r, eval=FALSE}
devtools::test_active_file()  # Ctrl+Shift+T in RStudio (for active file)
```

Or run all tests in the package:

```{r, eval=FALSE}
devtools::test()
```

Testing with OMOP data is covered in detail in @sec-test-omop. For now, the important habit is: every function you write should have at least one test file created for it before you move on.

## Create a README

The README is the first thing people see when they visit your package on GitHub. Create one with:

```{r, eval=FALSE}
use_readme_rmd()
```

This creates `README.Rmd`, which is an R Markdown file that can contain executable code chunks. After editing it, render it to produce `README.md`:

```{r, eval=FALSE}
devtools::build_readme()
```

Always commit both `README.Rmd` and `README.md`. GitHub displays `README.md` automatically on the repository front page.

A good README for an ecosystem package should include: a one-paragraph description, an installation code block, a minimal usage example using a `mock*` function (no real database needed), and links to the documentation website. The full README conventions are covered in @sec-readme.

## Create a documentation website

A [pkgdown](https://pkgdown.r-lib.org) website gives your package a searchable, browsable reference site generated automatically from your roxygen2 documentation and vignettes. Set it up with:

```{r, eval=FALSE}
use_pkgdown()
```

Build the site locally to preview it:

```{r, eval=FALSE}
pkgdown::build_site()
```

This creates a `docs/` folder. Open `docs/index.html` in your browser to see it.

The ecosystem uses a shared pkgdown theme for visual consistency. Configuration and automated deployment via GitHub Actions are covered in @sec-automatic-website.

## Set up GitHub

Once your package is working locally, push it to GitHub so it can benefit from continuous integration: automated checks, test coverage reporting, and website deployment on every push.

The fifth part of this book covers GitHub and GitHub Actions in full:

-   [Automatic checking your package](../github/automatic_checks.qmd)
-   [Automatic deployment of a documentation website](../github/automatic_website.qmd)
-   [Automatic assessment of test coverage](../github/test_coverage.qmd)
-   [Testing across multiple database backends](../github/testing_multiple_dbms.qmd)

## Using the package template

Although it is instructive to follow these steps manually the first time, the ecosystem provides a GitHub template repository that does all of this setup for you. Use it to create a new repository with the full scaffold already in place — license, GitHub Actions, pkgdown configuration, and lintr — by clicking the link below:

👉 [Create a new package from the template](https://github.com/new?template_name=EmptyPackageTemplate&template_owner=oxford-pharmacoepi)

::: {.callout-note collapse="true"}
### Template repository

The source for the template is at <https://github.com/oxford-pharmacoepi/EmptyPackageTemplate>. You can inspect it to see exactly what files are included.
:::
