# README

The `README.md` is a very important of an R package, is probably the first thing the users will read once they decide to try out our package. Now we will so how to generate it and how to populate it.

## Generate from `.Rmd`

In general we will populate our `README.md` using a rmarkdown file (`README.Rmd`) if you don't have it created you can create it with:

```{r, eval=FALSE}
library(usethis)
use_readme_rmd()
```

this will create the `README.Rmd` that generates the `README.md` file. To later generate the `README.md` you can use the following command:

```{r, eval=FALSE}
library(devtools)
build_readme()
```

If you are new to *rmakdown* probable this introduction will be useful: [Introduction to R Markdown](https://rmarkdown.rstudio.com/articles_intro.html).

::: {.callout-note collapse="true"}
### `README.md` and `README.Rmd`
Note that for cran only the `README.md` exists, so any change that you do to the `README.Rmd` file wont affect the 'real' README.md file till you build it, so make sure to build the `.md` file every time that you modify the `.Rmd`.
:::

## Content

### Badges

### Tested sources

### Goal of your package

### How to install

Make sure to include 

### Simple examples

Include diagrams, figures and tables. 
