# Core dependencies {#sec-core-dependencies}

The *Tidy R OMOP CDM* packages rely on [dplyr](https://dplyr.tidyverse.org) and [dbplyr](https://dbplyr.tidyverse.org) to manipulate data from a `cdm_reference` object. The `cdm_reference` is the central object in the ecosystem — a named list of lazy database table references, all pointing to the same underlying data source. It is defined in [omopgenerics](https://darwin-eu.github.io/omopgenerics/), the package that every other package in the ecosystem depends on.

This chapter describes what `omopgenerics` provides and which other core packages are useful enough to consider importing or suggesting.

## omopgenerics

`omopgenerics` defines the classes and methods that give the ecosystem its coherence. It is a deliberately minimal package: its only external dependencies are `dplyr`, `dbplyr`, `cli`, `glue`, and `rlang`. This keeps its footprint small so that depending on it does not burden downstream packages.

### Classes

The four main classes defined by `omopgenerics` are:

**`cdm_reference`** is the central object of the whole ecosystem. It is a named list whose elements are lazy table references pointing to OMOP CDM tables in a database, along with any cohort or other tables associated with the CDM. The `cdm_reference` knows which database back-end it is connected to, what the CDM name is, and what schema to write results into. Every analytical function in the ecosystem takes a `cdm_reference` (or a table derived from one) as its first argument.

```{r, eval=FALSE}
library(omopgenerics)

# Access CDM properties
cdmName(cdm)
cdmVersion(cdm)
cdmSource(cdm)
sourceType(cdm)

# Insert, read, list, and drop tables in the write schema
cdm <- insertTable(cdm, name = "my_table", table = my_data)
cdm <- readSourceTable(cdm, name = "my_table")
listSourceTables(cdm)
cdm <- dropSourceTable(cdm, name = "my_table")
```

**`cdm_table`** is the class of any table that belongs to a `cdm_reference`. All CDM tables carry a back-reference to their parent `cdm_reference`, which means you can always navigate from a table back to the full CDM:

```{r, eval=FALSE}
cdmReference(cdm$person)   # returns the full cdm_reference
cdmName(cdm$person)        # works just like cdmName(cdm)
tableName(cdm$person)      # returns "person"
```

**`omop_table`** is a subclass of `cdm_table` that protects the integrity of the standard OMOP CDM tables. It prevents removal of required columns and enforces that a table assigned to `cdm$person` must actually be named `person`. Filtering, mutating, and adding columns are all permitted — only operations that would break the CDM schema are blocked.

```{r, eval=FALSE}
# This is blocked — person_id is required
cdm$person <- cdm$person |> select(-person_id)
#> Error: person_id is not present in table person

# This is fine — adding a column is allowed
cdm$person <- cdm$person |> mutate(extra_col = "my_content")
```

**`cohort_table`** extends `cdm_table` with the structure required for a study cohort. A cohort table has exactly four columns (`cohort_definition_id`, `subject_id`, `cohort_start_date`, `cohort_end_date`) and carries three mandatory attributes stored as database tables: the cohort set (metadata per cohort), the attrition log (a record of how many records and subjects were retained at each step), and the cohort codelist (the clinical codes used to define each cohort).

```{r, eval=FALSE}
settings(cdm$my_cohort)      # cohort set: name, cdm_version, vocabulary_version
cohortCount(cdm$my_cohort)   # records and subjects per cohort
attrition(cdm$my_cohort)     # full attrition table

# Record a new attrition step after filtering
cdm$my_cohort <- cdm$my_cohort |>
  filter(subject_id %% 2 == 0) |>
  compute(name = "my_cohort") |>
  recordCohortAttrition("Only even subjects")
```

### Codelists

`omopgenerics` defines three classes for representing clinical code lists:

- **`codelist`**: a named list of integer vectors, one per concept set.
- **`codelist_with_details`**: a named list of tibbles, where each tibble contains `concept_id` alongside additional properties (e.g. `concept_name`, `domain_id`).
- **`concept_set_expression`**: a named list of tibbles with `concept_id`, `excluded`, `descendants`, and `mapped` columns — matching the ATLAS JSON format and the OHDSI TAB guidelines.

```{r, eval=FALSE}
# Create a simple codelist
cl <- newCodelist(list(hypertension = c(316866L, 320128L), diabetes = c(201826L)))

# Combine and subset
c(cl, newCodelist(list(asthma = 317009L)))
cl["hypertension"]

# Export and import (JSON is compatible with ATLAS)
exportCodelist(cl, path = here::here(), type = "json")
cl <- importCodelist(path = here::here(), type = "json")
```

In general, analytical functions in the ecosystem accept a `conceptSet` argument that can be any of these three classes (or a plain named list of integers that will be coerced to a `codelist`).

### The `summarised_result` class

`summarised_result` is the standard result format for all analytics functions in the ecosystem. It is a tibble with 13 compulsory columns. See @sec-function-outputs for a full description and guidance on how to create and populate it.

### Methods

`omopgenerics` exports a set of generic S3 methods that packages implement for their own classes:

| Generic | Purpose |
|---|---|
| `cdmName()` | Return the name of the CDM instance |
| `cdmVersion()` | Return the CDM version |
| `cdmSource()` | Return the `cdm_source` object |
| `sourceType()` | Return the back-end type (e.g. `"duckdb"`) |
| `tableName()` | Return the name of a `cdm_table` |
| `settings()` | Return cohort set metadata |
| `cohortCount()` | Return record/subject counts per cohort |
| `attrition()` | Return the attrition log |
| `cohortCodelist()` | Return the codelist used to build a cohort |
| `recordCohortAttrition()` | Append a step to the attrition log |
| `bind()` | Bind two `summarised_result` objects together |
| `suppress()` | Apply small-number suppression to a `summarised_result` |

### Input validation functions

`omopgenerics` provides `validate*` functions for the arguments that appear repeatedly across the ecosystem. Using these instead of writing bespoke validation code keeps error messages consistent across packages. See @sec-error-and-validation for full details and examples. The available validators are:

`validateCdmArgument()`, `validateCohortArgument()`, `validateCohortIdArgument()`, `validateConceptSetArgument()`, `validateNameArgument()`, `validateNameStyle()`, `validateResultArgument()`, `validateStrataArgument()`, `validateWindowArgument()`, `validateAgeGroupArgument()`.

There are also `assert*` helpers for simpler type checks: `assertCharacter()`, `assertNumeric()`, `assertLogical()`, `assertDate()`, `assertList()`, `assertClass()`.

### Manipulating a `cdm_reference`

In addition to the methods above, `omopgenerics` provides utilities for working with CDM table collections:

```{r, eval=FALSE}
# List tables by type
omopTables(cdm)       # standard OMOP CDM tables
cohortTables(cdm)     # cohort tables in the write schema
achillesTables(cdm)   # Achilles result tables

# Insert and drop
cdm <- insertTable(cdm, name = "temp", table = my_tibble)
cdm <- dropSourceTable(cdm, name = "temp")
```

### Manipulating a `summarised_result`

A `summarised_result` object has its own set of helper functions:

```{r, eval=FALSE}
# Inspect
settings(result)          # result-level metadata (result_type, package_name, ...)
groupColumns(result)      # columns encoded in group_name / group_level
strataColumns(result)     # columns encoded in strata_name / strata_level

# Transform
tidy(result)              # pivot to wide format — useful for internal computation
bind(result1, result2)    # combine two summarised_result objects
suppress(result, minCellCount = 5)  # apply cell suppression
```

## Other useful packages

Beyond `omopgenerics`, several other core packages are commonly useful. Unlike `omopgenerics`, these should generally be listed under `Suggests` unless your package's primary purpose depends on them.

**[PatientProfiles](https://darwin-eu.github.io/PatientProfiles/)** adds columns to cohort tables: demographics (age, sex, prior observation, future observation), intersection flags and values (e.g. "was the patient on drug X at cohort start?"), and date-relative summaries. Browse the [function reference](https://darwin-eu.github.io/PatientProfiles/reference/index.html) before implementing any feature-derivation logic — it is very likely already there.

**[visOmopResults](https://darwin-eu.github.io/visOmopResults/)** transforms `summarised_result` objects into formatted tables and `ggplot2` plots. The package supports multiple output formats for tables (gt, flextable, tibble) with a consistent API. See @sec-visualisations for guidance on how to use it in your package. Browse the [function reference](https://darwin-eu.github.io/visOmopResults/reference/index.html) before adding any result-formatting code.

**[CodelistGenerator](https://darwin-eu.github.io/CodelistGenerator/)** generates codelists by searching the OMOP vocabulary, and provides utilities for subsetting, stratifying, comparing, and diagnosing codelists. If your package works with clinical concepts, check the [function reference](https://darwin-eu.github.io/CodelistGenerator/reference/index.html) first.

**[omock](https://ohdsi.github.io/omock/)** creates synthetic OMOP CDM datasets for testing. It should almost always be listed under `Suggests`, not `Imports`. See @sec-test-omop for detailed guidance on using it in your test suite.

## Imports vs Suggests

A package listed under `Imports` is loaded automatically when your package is loaded. A package listed under `Suggests` is not — it must be available but is only used in specific circumstances (tests, vignettes, or optional code paths guarded by `requireNamespace()`).

The key rule for this ecosystem:

- **Always `Imports`**: `omopgenerics`, any package whose classes or generics you use in exported functions.
- **Usually `Suggests`**: `omock` (tests only), `CDMConnector` / back-end packages (tests only), `visOmopResults` (unless your package's core output is visualisation), `PatientProfiles` (unless your package is built on top of it).

When in doubt, prefer `Suggests`. A lighter `Imports` list means fewer forced transitive dependencies for your users.
