# Error messages and input validation {#sec-error-and-validation}

Giving users informative error messages is one of the most important things you can do to make a package pleasant to use. A cryptic error from deep inside a dplyr pipeline — or worse, a silent wrong result — is much harder to debug than a clear message at the top of the function that says exactly what was wrong with the input.

The rule of thumb is: **validate at the boundary, trust inside**. Check all inputs at the very start of each exported function, before any work is done. Once those checks pass, do not re-check the same things in internal helpers.

That said, do not overdo it. Avoid checking things that are guaranteed by earlier checks, and avoid checks that cost more than the function itself. In general, check that each argument has the expected type, that it has the expected length, and that any logical constraints between arguments are satisfied.

## The two kinds of validation helpers in `omopgenerics`

`omopgenerics` provides two families of validation helpers that together cover almost everything you need.

**`validate*` functions** are for arguments that appear consistently across many packages — `cdm`, `cohort`, `cohortId`, `conceptSet`, `window`, `ageGroup`, `strata`, `name`, `nameStyle`, and `result`. These functions do more than just check: they also **coerce and normalise** the input to a canonical form. You must assign their return value back to the variable, because the cleaned-up version is what the rest of your function should use.

**`assert*` functions** are for simple type and constraint checks. They do not transform the input — they either pass silently or throw an error. Use these for arguments that do not have a dedicated `validate*` function.

## `validate*` functions

The following `validate*` functions are available in `omopgenerics`. Import them with `@importFrom omopgenerics validateCdmArgument` (and so on) rather than calling them with `omopgenerics::`.

| Function | Validates and normalises |
|---|---|
| `validateCdmArgument(cdm)` | CDM reference: checks class, optionally checks required tables |
| `validateCohortArgument(cohort)` | Cohort table: checks class, required columns, attributes |
| `validateCohortIdArgument(cohortId, cohort)` | Cohort IDs: resolves `NULL` to all IDs, checks IDs exist in cohort |
| `validateConceptSetArgument(conceptSet)` | Concept set: accepts codelist, codelist_with_details, or concept_set_expression |
| `validateWindowArgument(window)` | Window: accepts vector or named list, always returns named list |
| `validateAgeGroupArgument(ageGroup)` | Age groups: accepts several input forms, always returns named list of named intervals |
| `validateStrataArgument(strata, cohort)` | Strata: checks that all columns named in strata exist in the cohort |
| `validateNameArgument(name, cdm)` | Table name: checks it is a valid identifier, optionally checks it doesn't already exist |
| `validateNameStyle(nameStyle, ...)` | Name style template: checks it is a valid glue template |
| `validateResultArgument(result)` | Summarised result: checks class and required columns |

### Normalisation is the point

The most important property of `validate*` functions is that they accept several reasonable input forms and always return a single canonical form. This means the rest of your function only has to handle one form, which keeps the logic clean.

The `ageGroup` argument illustrates this well. Users can supply it in several ways:

```{r}
library(omopgenerics, warn.conflicts = FALSE)

# Simple unnamed intervals
ageGroup <- validateAgeGroupArgument(ageGroup = list(c(0, 17), c(18, 64)))
ageGroup

# Complex named multi-column form
ageGroup <- validateAgeGroupArgument(ageGroup = list(
  age_at_start = list("child" = c(0, 17), "adult" = c(18, 64)),
  age_at_end   = list(c(0, 49), c(50, Inf))
))
ageGroup
```

Whatever you pass in, the output is always a named list of named intervals. The function also throws clear errors if the input is malformed:

```{r, error=TRUE}
# Overlapping age groups when only one column is allowed
validateAgeGroupArgument(
  ageGroup = list(group1 = list(c(0, 19), c(20, Inf)), group2 = list(c(0, Inf))),
  multipleAgeGroup = FALSE
)

# Negative ages
validateAgeGroupArgument(ageGroup = list(c(-5, 19), c(20, Inf)))

# NULL not allowed
validateAgeGroupArgument(ageGroup = NULL, null = FALSE)
```

The same principle applies to `validateWindowArgument()`, which accepts either a plain `c(0, Inf)` vector or a named list of windows, and always returns a named list.

### The `{{` syntax for `cohortId`

`validateCohortIdArgument()` supports [tidyselect](https://tidyselect.r-lib.org) semantics, so users can pass things like `starts_with("fracture")` to select cohort IDs by name. To make this work, you must wrap the argument in double curly braces `{{` when you pass it to the validator:

```{r}
#| eval: false
cohortId <- validateCohortIdArgument(cohortId = {{cohortId}}, cohort = cohort)
```

Without `{{`, tidyselect expressions will not be evaluated correctly.

## `assert*` functions

Use `assert*` functions for arguments that do not have a dedicated `validate*` function. Unlike `validate*`, these functions return the input invisibly and are used purely for their side effect of throwing an error.

| Function | Checks |
|---|---|
| `assertCharacter(x, ...)` | `x` is a character vector |
| `assertNumeric(x, ...)` | `x` is numeric |
| `assertLogical(x, ...)` | `x` is logical |
| `assertDate(x, ...)` | `x` is a Date |
| `assertList(x, ...)` | `x` is a list |
| `assertClass(x, class, ...)` | `x` has the specified class(es) |
| `assertChoice(x, choices, ...)` | `x` is one of the allowed values |
| `assertTable(x, ...)` | `x` is a table (data frame or `tbl_sql`) |
| `assertTrue(expr, ...)` | The expression evaluates to `TRUE` |

All of these accept common modifiers as arguments:

| Modifier | Type | Meaning |
|---|---|---|
| `length` | `integer` or `NULL` | Required length; `NULL` skips the check |
| `null` | `logical` | Whether `NULL` is a valid input (default `FALSE`) |
| `na` | `logical` | Whether `NA` values are allowed (default `FALSE`) |
| `named` | `logical` | Whether elements must be named |
| `unique` | `logical` | Whether elements must be unique |
| `min` | `numeric` | Minimum value (for `assertNumeric`) |
| `max` | `numeric` | Maximum value (for `assertNumeric`) |
| `integerish` | `logical` | Whether numeric must be whole numbers (for `assertNumeric`) |
| `minNchar` | `integer` | Minimum number of characters per element (for `assertCharacter`) |
| `call` | call | Passed to the `cli` error message for better call context |

## Putting it together: a validation block

Here is the recommended pattern. Validation comes first, before any computation. `validate*` results are always reassigned. `assert*` calls are written as statements.

```{r}
myFunction <- function(cohort, cohortId = NULL, window = c(0, Inf), overlap = FALSE) {
  # validate and normalise
  cohort   <- omopgenerics::validateCohortArgument(cohort = cohort)
  cohortId <- omopgenerics::validateCohortIdArgument(cohortId = {{cohortId}}, cohort = cohort)
  window   <- omopgenerics::validateWindowArgument(window = window)
  
  # simple type checks
  omopgenerics::assertLogical(overlap, length = 1)
  
  # business logic checks
  if (overlap && any(sapply(window, function(w) diff(w) > 365))) {
    cli::cli_abort(c(
      "x" = "Windows longer than 365 days are not supported when {.var overlap} is {.val TRUE}."
    ))
  }
  
  # ... function body
}
```

A second example with a CDM reference:

```{r}
myFunction2 <- function(cdm, conceptSet, days = 180L, startDate = NULL, overlap = TRUE) {
  # validate and normalise
  cdm        <- omopgenerics::validateCdmArgument(cdm = cdm)
  conceptSet <- omopgenerics::validateConceptSetArgument(conceptSet = conceptSet)
  
  # simple type checks
  omopgenerics::assertNumeric(days, integerish = TRUE, min = 0, length = 1)
  omopgenerics::assertDate(startDate, length = 1, null = TRUE)
  omopgenerics::assertLogical(overlap, length = 1)
  
  # business logic checks
  if (overlap && days > 365) {
    cli::cli_abort(c(
      "x" = "{.var days} cannot be >= 365 when {.var overlap} is {.val TRUE}."
    ))
  }
  
  # ... function body
}
```

## Writing good error messages with `cli`

Custom errors and warnings should use the [**cli**](https://cli.r-lib.org) package, which is already a dependency of `omopgenerics`. Do not use `stop()` or `warning()`.

```{r}
#| eval: false
# Error
cli::cli_abort(c(
  "x" = "Argument {.var days} must be a positive integer.",
  "i" = "You supplied {.val {days}}."
))

# Warning  
cli::cli_warn(c(
  "!" = "Cohort {.val {name}} already exists in the CDM and will be overwritten."
))

# Informational message
cli::cli_inform(c(
  "i" = "Computing {length(cohortId)} cohort{?s}."
))
```

A few conventions for error messages:

- Use `{.var argument_name}` to refer to an argument by name.
- Use `{.val {value}}` to show the actual value that was supplied.
- Use `{.fn function_name}` to refer to a function.
- Use `{.cls class_name}` to refer to a class.
- Use named elements `"x"` for errors, `"!"` for warnings, `"i"` for hints or additional information.
- Write error messages in the present tense from the user's perspective: "must be", "cannot be", "is not" rather than "expected", "got".

## What to validate and what to skip

Validate arguments that users are likely to pass in many ways or get wrong:

- Arguments with a standard `validate*` function — always validate these.
- Type-sensitive arguments: anything the rest of the function passes directly to dplyr or database operations.
- Arguments with constraints relative to each other (e.g. `days` must be positive when `overlap` is `TRUE`).

Do not validate:

- Internal function arguments that are never user-facing.
- Things already guaranteed by a `validate*` call (e.g. do not re-check that `cohortId` is numeric after calling `validateCohortIdArgument`).
- Constants you define yourself inside the function.

## Summary

Validating arguments is one of the most impactful things you can do for users — it turns opaque database errors into clear messages that point directly to the problem. `omopgenerics` makes this straightforward: use `validate*` for the standard OMOP arguments (and always reassign the result), use `assert*` for simple type checks, and use `cli` for any custom messages. A complete validation block for a typical function takes four or five lines and requires no custom logic.
