# Function outputs {#sec-function-outputs}

## The `summarised_result` object

The `summarised_result` is the standard result format returned by all analytics functions in the ecosystem. It is an S3 class built on top of a tibble, defined in `omopgenerics`. Every package that produces estimates — prevalence, drug use, cohort characteristics, survival — returns a `summarised_result`. This standardisation is what allows `visOmopResults` and `OmopViewer` to work uniformly across the whole ecosystem without any package-specific code.

### Structure

A `summarised_result` is a tibble with exactly 13 compulsory columns:

| Column | Type | NA allowed | Description |
|---|---|---|---|
| `result_id` | integer | No | Groups rows sharing the same result settings. |
| `cdm_name` | character | No | Name of the CDM instance used. |
| `group_name` | character | No | Name side of the group name–level pair. |
| `group_level` | character | No | Level side of the group name–level pair. |
| `strata_name` | character | No | Name side of the strata name–level pair. |
| `strata_level` | character | No | Level side of the strata name–level pair. |
| `variable_name` | character | No | Name of the variable being estimated. |
| `variable_level` | character | Yes | Sub-level of the variable (e.g. a category). |
| `estimate_name` | character | No | Name of the estimate (snake_case). |
| `estimate_type` | character | No | Type of the estimate value (see below). |
| `estimate_value` | character | No | Value of the estimate, always stored as character. |
| `additional_name` | character | No | Name side of the additional name–level pair. |
| `additional_level` | character | No | Level side of the additional name–level pair. |

The `estimate_type` column must be one of the values returned by `omopgenerics::estimateTypeChoices()`: `"numeric"`, `"integer"`, `"date"`, `"character"`, `"proportion"`, `"percentage"`, `"logical"`.

#### Name–level pairs

Three pairs of columns — `group_name`/`group_level`, `strata_name`/`strata_level`, `additional_name`/`additional_level` — follow the same encoding. When a result applies to multiple dimensions simultaneously, the column names are concatenated with ` &&& ` and the values likewise:

```
group_name:  "cohort_name &&& sex"
group_level: "asthma_cohort &&& Female"
```

- **group**: high-level identifiers, typically cohort name or codelist name.
- **strata**: stratification dimensions such as age group or sex.
- **additional**: any further dimensions that don't fit in group or strata (e.g. time window, analysis type).

When a result is not stratified, these columns should contain `"overall"` / `"overall"`.

### Settings attribute

Every `summarised_result` has a `settings` attribute — a tibble with one row per `result_id` — that stores metadata about how the results were generated. Three settings columns are compulsory:

| Setting | Description |
|---|---|
| `result_type` | A short snake_case string identifying the type of analysis (e.g. `"summarised_characteristics"`). Must be unique within a package. |
| `package_name` | The name of the package that generated the result. |
| `package_version` | The version of that package. |

Additional settings columns store analysis-specific parameters (e.g. `strata`, `ageGroup`, `window`). A `min_cell_count` column is always present, defaulting to `"0"`.

### How to create a `summarised_result`

The standard pattern is to build a plain tibble with the 13 required columns, then convert it using `omopgenerics::newSummarisedResult()`:

```{r, eval=FALSE}
library(omopgenerics)

result <- tibble::tibble(
  result_id       = 1L,
  cdm_name        = cdmName(cdm),
  group_name      = "cohort_name",
  group_level     = "my_cohort",
  strata_name     = "overall",
  strata_level    = "overall",
  variable_name   = "age",
  variable_level  = NA_character_,
  estimate_name   = "mean",
  estimate_type   = "numeric",
  estimate_value  = as.character(mean_age),
  additional_name = "overall",
  additional_level= "overall"
) |>
  newSummarisedResult(
    settings = tibble::tibble(
      result_id       = 1L,
      result_type     = "my_result_type",
      package_name    = "myPackage",
      package_version = as.character(packageVersion("myPackage"))
    )
  )
```

In practice you will produce many rows at once. The recommended internal workflow is:

1. Compute a tidy intermediate tibble (one row per subject / cohort / stratum / estimate).
2. Use `visOmopResults::uniteNameLevel()` or `omopgenerics` helpers to encode name–level pairs.
3. Call `newSummarisedResult()` with the appropriate settings.

### Using the tidy format internally

The `tidy()` method converts a `summarised_result` back to a wide, human-readable format:

```{r, eval=FALSE}
result |> tidy()
```

This is useful for inspecting results during development, and can also be used internally if your function needs to manipulate results before returning them. However, always return a `summarised_result` from exported functions — never a tidy tibble.

### Suppression

If your function produces counts of patients or records, it must support cell suppression. The convention is to expose a `minCellCount` argument (default `5`) and apply suppression before returning:

```{r, eval=FALSE}
result <- result |> suppress(minCellCount = minCellCount)
```

`suppress()` replaces estimate values with `NA` where the count is below the threshold, and records the suppression threshold in the settings.

If your result contains a count of subjects (i.e. unique `subject_id` values), the estimate should be named `count_subjects` or `n_subjects` so that suppression is applied correctly.

### One `result_type` per function

Each exported function that returns a `summarised_result` should produce exactly one `result_type`. Do not mix different analyses in a single `summarised_result` returned from one function — this makes it impossible for downstream tools to filter and process results by type.

### Registering a `result_type`

Result types should be registered in your package so that `visOmopResults` and `OmopViewer` can discover them. Add a call to `omopgenerics::registerResultType()` in your package's `.onLoad()` or document it in the package metadata. See the `omopgenerics` documentation for the current registration API.

## Cohort outputs

Functions that return a modified cohort (rather than a summary) should return a `cohort_table`. The attrition log must be updated for every filtering step using `recordCohortAttrition()`. See @sec-core-dependencies for the `cohort_table` class description.

## Further reading

- [`omopgenerics` summarised_result documentation](https://darwin-eu.github.io/omopgenerics/articles/summarised_result.html)
- [`visOmopResults` package](https://darwin-eu.github.io/visOmopResults/)
