# Styling your code

Consistent code style across the ecosystem means that any contributor can read and navigate any package without having to mentally re-parse a different formatting convention. This chapter describes the style rules used and the tools that enforce them.

## Code style

The ecosystem follows the [tidyverse style guide](https://style.tidyverse.org) with one important modification: **function and argument names use camelCase**, not snake_case. This applies to all exported functions, unexported functions, and arguments. Object names inside function bodies should also use camelCase.

```{r, eval=FALSE}
# Correct
summariseCohortOverlap <- function(cohort, cohortId = NULL, minCellCount = 5) {
  resultData <- computeOverlap(cohort, cohortId)
  resultData |> formatOutput(minCellCount = minCellCount)
}

# Incorrect — snake_case names
summarise_cohort_overlap <- function(cohort, cohort_id = NULL, min_cell_count = 5) {
  result_data <- compute_overlap(cohort, cohort_id)
}
```

The rationale for camelCase over snake_case is historical consistency: the earliest packages in the ecosystem (including `omopgenerics` and `PatientProfiles`) used camelCase, and changing convention mid-ecosystem would introduce more confusion than it resolves.

## Linting

All packages in the ecosystem use [`lintr`](https://lintr.r-lib.org) to enforce style. The standard configuration is:

```{r, eval=FALSE}
lintr::lint_package(
  linters = lintr::linters_with_defaults(
    lintr::object_name_linter(styles = "camelCase")
  )
)
```

This checks all camelCase naming conventions, line length, spacing, and other tidyverse style rules, with the camelCase override for names.

You should run `lint_package()` before opening any pull request. The GitHub Actions CI workflow (see @sec-automatic-checks) runs linting automatically on every push.

To add lintr configuration to your package so that the linter settings are stored in the repository:

```{r, eval=FALSE}
usethis::use_lintr()
```

This creates a `.lintr` file at the root of your package. Edit it to match the ecosystem configuration:

```
linters: linters_with_defaults(
  object_name_linter(styles = "camelCase")
)
```

## Formatting

Use [`styler`](https://styler.r-lib.org) to automatically format code. The standard command is:

```{r, eval=FALSE}
styler::style_pkg()
```

::: {.callout-important}
When contributing to an existing package, **do not restyle code that you have not otherwise modified**. Reformatting unrelated lines makes diffs harder to review and is a common source of merge conflicts. Only restyle the lines you have actively changed.
:::

## Line length

The maximum line length is 80 characters. Long function calls should be broken across lines with one argument per line, with the closing parenthesis on its own line:

```{r, eval=FALSE}
# Correct
result <- summariseCharacteristics(
  cohort = cdm$my_cohort,
  strata = list(c("sex"), c("age_group")),
  minCellCount = 5
)

# Avoid — too long for a single line
result <- summariseCharacteristics(cohort = cdm$my_cohort, strata = list(c("sex"), c("age_group")), minCellCount = 5)
```

## Pipes

Use the native pipe `|>` introduced in R 4.1. The `%>%` pipe from magrittr should not be used in new code. The native pipe is imported automatically and requires no additional dependency.

## Assignment

Use `<-` for assignment, not `=`. Reserve `=` for argument passing inside function calls.

## Spacing

- Put spaces around all binary operators (`<-`, `+`, `-`, `==`, `|>`, etc.).
- Put a space after commas but not before.
- Do not put spaces inside parentheses: `f(x, y)` not `f( x, y )`.

These rules are enforced by `lintr` and will be flagged automatically.
