# Visualisations {#sec-visualisations}

All visualisation in the ecosystem is built on top of `visOmopResults`. This package provides a consistent API for transforming `summarised_result` objects into formatted tables and `ggplot2` plots. By using it rather than writing bespoke formatting code, your package inherits consistent output across all result types and avoids duplicating logic that is already maintained centrally.

This chapter explains the conventions for adding `table*` and `plot*` functions to your package.

## The `visOmopResults` API

`visOmopResults` exposes two main families of functions:

- **`visOmopTable()`** and its wrappers produce formatted tables. The output format (gt, flextable, tibble, or datatable) is controlled by a `type` argument.
- **`visOmopPlot()`** and specific plot helpers produce `ggplot2` objects that can be further customised.

Most packages do not call these low-level functions directly. Instead they implement thin `table*` and `plot*` wrapper functions that call into `visOmopResults` with sensible defaults.

## Adding `table*` functions

Every `summarise*` function should have a corresponding `table*` function that formats its output for display. The convention is:

```{r, eval=FALSE}
tableMyResult <- function(result, type = "gt", ...) {
  # optionally filter/reshape result before passing to visOmopResults
  visOmopResults::visOmopTable(
    result = result,
    type   = type,
    ...
  )
}
```

Key points:

- The first argument is always a `summarised_result`, typically the output of the matching `summarise*` function.
- `type` defaults to `"gt"` in function signatures but **use `type = "flextable"` in README examples** — flextable renders better in GitHub Markdown.
- Pass `...` through to `visOmopResults::visOmopTable()` to expose column renaming, formatting, and other options without duplicating them in your own function signature.
- Validate the `result` argument using `omopgenerics::validateResultArgument()`.

A minimal example:

```{r, eval=FALSE}
#' Format summarised characteristics as a table
#'
#' @param result A \code{summarised_result} object as returned by
#'   \code{summariseCharacteristics()}.
#' @param type Output format. One of \code{"gt"} (default), \code{"flextable"},
#'   \code{"tibble"}, or \code{"datatable"}.
#' @param ... Additional arguments passed to
#'   \code{\link[visOmopResults]{visOmopTable}}.
#'
#' @return A formatted table object of the class determined by \code{type}.
#' @export
tableCharacteristics <- function(result, type = "gt", ...) {
  result <- omopgenerics::validateResultArgument(result)
  visOmopResults::visOmopTable(result = result, type = type, ...)
}
```

## Adding `plot*` functions

Similarly, `plot*` functions wrap `visOmopResults` plot helpers:

```{r, eval=FALSE}
#' Plot incidence estimates
#'
#' @param result A \code{summarised_result} object as returned by
#'   \code{estimateIncidence()}.
#' @param ... Additional arguments passed to the underlying plot function.
#'
#' @return A \code{ggplot2} object.
#' @export
plotIncidence <- function(result, ...) {
  result <- omopgenerics::validateResultArgument(result)
  visOmopResults::plotEstimates(result = result, ...)
}
```

Because `plot*` functions return `ggplot2` objects, users can add layers, change themes, and save the plot using standard ggplot2 idioms. Do not apply a custom theme inside your package's plot functions — leave theming to the user.

## Checking what is already available

Before writing any visualisation code, check the [visOmopResults function reference](https://darwin-eu.github.io/visOmopResults/reference/index.html). The package is actively extended, and the function you need may already exist. Duplicating it in your own package creates a maintenance burden and means your users get a subtly different experience.

## What not to do

- Do not return raw `gt` or `flextable` objects directly from `table*` functions — always go through `visOmopResults::visOmopTable()` so that the `type` argument works consistently.
- Do not hard-code column names or formatting inside your package — use the `visOmopResults` API so that changes to the formatting logic can be made in one place.
- Do not depend on `ggplot2` directly unless your package needs to construct plots from scratch (i.e. not from a `summarised_result`). For `summarised_result`-based plots, use `visOmopResults`.

## Suggesting vs importing `visOmopResults`

For most packages, `visOmopResults` should be listed under `Imports` rather than `Suggests`, because the `table*` and `plot*` functions are part of the package's exported API and will fail if `visOmopResults` is not installed. If your package only uses `visOmopResults` in vignettes or optional code paths, `Suggests` is appropriate.
